---
title: "Custom analyses Bjorkman mAbs"
author: "Allie Greaney"
date: "1/11/2021"
output:
  github_document:
    toc: true
    html_preview: false
editor_options: 
  chunk_output_type: inline
---

This notebook does some random analyses on the clinical antibodies set that vary from the more constrained global pipeline.

```{r setup, message=FALSE, warning=FALSE, error=FALSE}
require("knitr")
knitr::opts_chunk$set(echo = T)
knitr::opts_chunk$set(dev.args = list(png = list(type = "cairo")))

#list of packages to install/load
packages = c("yaml","data.table","tidyverse","ggrepel","bio3d")
#install any packages not already installed
installed_packages <- packages %in% rownames(installed.packages())
if(any(installed_packages == F)){
  install.packages(packages[!installed_packages])
}
#load packages
invisible(lapply(packages, library, character.only=T))

#read in config file
config <- read_yaml("config.yaml")

#read in escape profiles file
profiles_config <- read_yaml(file=config$escape_profiles_config)

#read in structural contacts config
structural_contacts_config <- read_yaml(file=config$structural_contacts_config)

#make output directory
output_dir <- './results/bjorkman_custom_analyses'
if(!file.exists(output_dir)){
  dir.create(file.path(output_dir))
}
```
Session info for reproducing environment:
```{r print_sessionInfo}
sessionInfo()
```

## Data input and formatting

Read in escape fractions, rename some things to make them easier to work with.

```{r data input}
#read in strong escape sites, keep sensitive threshold
strong_escape_sites <- data.table(read.csv(file=config$strong_escape_sites, stringsAsFactors=F)) %>%
  mutate(site_threshold=pmax(exceed_median, frac_max)) %>%
  select(condition, site, threshold, site_threshold) %>%
  filter(threshold=="sensitive") %>%
  mutate(strong_escape_site=TRUE)

head(strong_escape_sites)

# read in escape fractions
# scores <- scores[selection %in% names(profiles_config$Rockefeller_antibodies$conditions) & library=="average", .(selection,condition,site,protein_site,wildtype,mutation,mut_escape_frac_epistasis_model,site_total_escape_frac_epistasis_model)]

scores <- data.table(read.csv(file=config$escape_fracs,stringsAsFactors=F)) %>%
  filter(selection %in% names(profiles_config$Rockefeller_antibodies$conditions) & 
           library=="average") %>%
  select(selection,
         condition,
         protein_site,
         wildtype,
         mutation,
         mut_escape_frac_epistasis_model,
         site_total_escape_frac_epistasis_model) %>%
  merge(strong_escape_sites %>% select(-site_threshold),
        by.x=c("condition", "protein_site"),
        by.y=c("condition", "site"),
        all.x = TRUE) %>%
  replace_na(list(strong_escape_site=FALSE)) %>%
  merge(strong_escape_sites %>% select(condition, site_threshold) %>% unique(),
        by=c("condition")
        ) %>%
  rename(mut_escape_frac=mut_escape_frac_epistasis_model, 
         site_total_escape=site_total_escape_frac_epistasis_model
         )

# setnames(scores,"mut_escape_frac_epistasis_model","mut_escape_frac");setnames(scores,"site_total_escape_frac_epistasis_model","site_total_escape")
scores<-data.table(scores)
scores[,antibody:=as.character(profiles_config$Rockefeller_antibodies$conditions[condition])]

scores <- scores[,.(antibody,protein_site,wildtype,mutation,mut_escape_frac,site_total_escape,strong_escape_site, site_threshold)]

scores[,site_max_escape:=max(mut_escape_frac,na.rm=T),by=c("antibody","protein_site")]

scores <- scores %>% rename(site=protein_site)
head(scores%>%filter(strong_escape_site==FALSE))
```

## Part 1: antibody escape vs. proximity to the structural epitope

We want to try a few plots -- at the per-mut and per-site level, compare distribution of escape scores between structural contacts and non-contacts. We should try the contact/non-contact, but also a "close" category corresponding to 4-8A distance? 

Means we'll probably want to re-do the structural analysis here within this "custom analyses" script.

First, load pdbs, and generate list of residues that are <4A or 4-8A from each mAb

```{r define_structural_contacts}
#empty data frame to append contacts to
structural_contacts <- data.frame(antibody=character(0), site=numeric(0), site_category=character(0))

bjorkman_mabs <- profiles_config$Rockefeller_antibodies$conditions

RBD_sites <- c(333:526)

for (antibody in bjorkman_mabs){
  # make empty vectors to put contact sites in 
  contacts_4A <- c()
  contacts_8A <- c()
  unresolved <- c()
  
  for(entry in contacts_config){
    if ((unlist(strsplit(entry$name,'_'))[1] == antibody)){
      print(antibody)
      pdb <- read.pdb(file=entry$pdbfile)
      pdb_contacts_4A <- binding.site(pdb,
                               a.inds=atom.select(pdb,chain=entry$chains_RBD),
                               b.inds=atom.select(pdb,chain=entry$chains_ligand),
                               cutoff=4, hydrogens=F)$resno %>% unique()
      pdb_contacts_8A <- binding.site(pdb,
                               a.inds=atom.select(pdb,chain=entry$chains_RBD),
                               b.inds=atom.select(pdb,chain=entry$chains_ligand),
                               cutoff=8, hydrogens=F)$resno %>% unique()
      
      # there should be as many instances of each residue as there are RBD chains
      n_chains <- entry$chains_RBD %>% length()
      inds <- atom.select(pdb,chain=entry$chains_RBD, elety = c("CA"), resno=RBD_sites) #$resno #%>% unique()
      resi_count <- as.data.frame(table(pdb$atom[inds$atom,]$resno))
      colnames(resi_count) <- c("site", "Freq")
      resi_count$site <- as.numeric(levels(resi_count$site))[resi_count$site]
      print(resi_count)
      # merge with all integers from 333 to 526 to account for sites that are missing from all chains 
      if (length(resi_count)>0){
        resi_count <- resi_count %>% merge(data.frame(site=RBD_sites), by='site', all=TRUE)
        resi_count[is.na(resi_count)] <- 0
      }
      # resi_count <- resi_count %>% 
      # if there are fewer, then that residue is unresolved in >=1 RBD chain
      pdb_unresolved <- resi_count %>% filter(Freq < 2) %>% pull(site) %>% as.numeric()
      contacts_4A <- c(contacts_4A, pdb_contacts_4A)
      contacts_8A <- c(contacts_8A, pdb_contacts_8A)
      unresolved <- c(unresolved, pdb_unresolved)
    }
  }
  contacts_4A <- contacts_4A %>% unique()
  contacts_8A <- contacts_8A %>% unique()
  contacts_4Ato8A <- contacts_8A[!(contacts_8A %in% contacts_4A)]
  # if an unresolved site is called as a contact site in any other chain, exclude
  unresolved <- unresolved[!(unresolved %in% contacts_4A | unresolved %in% contacts_4Ato8A)] %>% unique()
  print('these are unresolved sites that are not contact sites:')
  print(unresolved)
  # now add to data frame 
  structural_contacts <- rbind(structural_contacts, 
                               data.frame(antibody=antibody,
                                          site = contacts_4A) %>% 
                                          mutate(site_category="contact")
                               )
  structural_contacts <- rbind(structural_contacts, 
                               data.frame(antibody=antibody,
                                          site = contacts_4Ato8A) %>% 
                                          mutate(site_category="proximal")
                               )
  if (length(unresolved)>0){
  structural_contacts <- rbind(structural_contacts, 
                               data.frame(antibody=antibody,
                                          site = unresolved) %>% 
                                          mutate(site_category="unresolved")
  )
  }
}

structural_contacts
```
```{r merge_scores_with_contacts}

structural_contacts$site <- as.integer(structural_contacts$site)

site_dataframe <- scores %>% 
  select(-mutation, -mut_escape_frac) %>% 
  unique() %>%
  merge(structural_contacts, by=c("antibody", "site"), all.x = TRUE) %>%
  replace_na(list(site_category="distal"))

site_dataframe
write.csv(site_dataframe, file=paste(output_dir,"/site_contact_escape.csv",sep=""), row.names = FALSE, quote = FALSE)
```

Table with number of sites in each structural category, and number of sites of "strong" escape

```{r get_labels_for_plot}
site_dataframe_labels <- site_dataframe %>% 
  group_by(antibody, strong_escape_site, site_category) %>% 
  count() %>%
  rename(n_sites=n) %>%
  pivot_wider(#id_cols = c(antibody, site_category),
    names_from = strong_escape_site,
    names_prefix = "strong_",
    names_repair = "check_unique",
    values_from = n_sites) %>%
  replace_na(list(strong_TRUE=0)) %>%
  mutate(sites_in_dist_cat=strong_FALSE+strong_TRUE,
         cat_label = paste0(strong_TRUE, "/", sites_in_dist_cat)
         )

site_dataframe_labels
```

Next, make jitterplots showing site total escape for the three categories of structural class.

```{r plot_total_escape_contacts, echo=T, fig.width=6, fig.height=3, fig.align="center", dpi=300,dev="png"}
site_dataframe$site_category <- factor(site_dataframe$site_category, 
                                       levels=c("contact", "proximal", "distal", "unresolved"))
site_dataframe$antibody <- factor(site_dataframe$antibody, 
                                  levels=c("C105", "C144", "C002", "C121", "C135", "C110"))
site_dataframe$strong_escape_site <- factor(site_dataframe$strong_escape_site, 
                                  levels=c(TRUE, FALSE))

site_dataframe_labels$antibody <- factor(site_dataframe_labels$antibody, 
                                  levels=c("C105", "C144", "C002", "C121", "C135", "C110")) 
  
ggplot(site_dataframe, 
       aes(x=site_category, y=site_total_escape, color=strong_escape_site)) +
  geom_jitter(width=0.15, alpha=0.5,height=0,size=2.25,shape=16) +
  geom_hline(data=site_dataframe %>% select(antibody,site_threshold) %>% unique(),
             aes(yintercept=site_threshold),
             linetype="dashed",
             color="#999999")+
  scale_y_continuous(limits=c(0, 22))+
  theme_classic()+
  theme(text = element_text(size=12), 
        axis.text.x=element_text(angle=90, hjust=1, vjust=0.5, size=12))+
  xlab("structural context")+
  ylab("total escape at site")+
  geom_text(data=site_dataframe_labels, 
            aes(x=site_category, y=21), 
            size=2.5,
            color="#D55E00",
            alpha=1,
            inherit.aes = FALSE,
            label=site_dataframe_labels$cat_label) +
  facet_wrap(~antibody,nrow=1) +
  scale_color_manual(values=c('#D55E00', '#999999'), name="site of strong\nescape")

ggsave(
  paste(output_dir,"/escape_by_contact_type.pdf",sep=""),
  plot = last_plot(),
  width = 9,
  height = 2.5,
)
```
I am curious about the C110 distal sites that are strong escape sites.
This Fab-spike trimer structure has a number of unresolved residues in the structural footprint.
I wonder if that's what those sites are. 
```{r investigate_certain_sites}
site_dataframe %>% filter(site_category=="distal" & strong_escape_site==TRUE)
```

